---
title: "BIS 15L Midterm 2"
author: "Qianyu Jin"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your code should be organized, clean, and run free from errors. Be sure to **add your name** to the author header above. You may use any resources to answer these questions (including each other), but you may not post questions to Open Stacks or external help sites. There are 10 total questions.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean! Your plots should use consistent aesthetics throughout.  

This exam is due by **12:00p on Tuesday, February 22**.  
```{r}
library(tidyverse)
library(janitor)
library(here)
library(naniar)
library("gapminder")
library(RColorBrewer)
library(paletteer)
```

## Gapminder
For this assignment, we are going to use data from  [gapminder](https://www.gapminder.org/). Gapminder includes information about economics, population, social issues, and life expectancy from countries all over the world. We will use three data sets, so please load all three as separate objects.    

1. population_total.csv  
2. income_per_person_gdppercapita_ppp_inflation_adjusted.csv  
3. life_expectancy_years.csv  
```{r}
population <- read_csv(here("midterm2", "data", "population_total.csv"))
income_per_person <- read_csv(here("midterm2", "data", "income_per_person_gdppercapita_ppp_inflation_adjusted.csv"))
life_expectancy <- read_csv(here("midterm2", "data", "life_expectancy_years.csv"))
```

1. (3 points) Once you have an idea of the structure of the data, please make each data set tidy (hint: think back to pivots) and store them as new objects. You will need both the original (wide) and tidy (long) data! 
```{r}
summary(population)
```


```{r}
summary(income_per_person)
```


```{r}
summary(life_expectancy)
```

```{r}
population_tidy <- population %>%
  pivot_longer(!country, 
               names_to = "year", 
               values_to = "population")
```


```{r}
income_per_person_tidy <- income_per_person %>%
  pivot_longer(!country, 
               names_to = "year", 
               values_to = "income")
```

```{r}
life_expectancy_tidy <- life_expectancy %>%
  pivot_longer(!country, 
               names_to = "year", 
               values_to = "life_expectancy")
```

2. (1 point) How many different countries are represented in the data? Provide the total number and their names. Since each data set includes different numbers of countries, you will need to do this for each one.  
```{r}
population_tidy %>%
  summarise(total_contries = n_distinct(country))
```
```{r}
income_per_person_tidy %>%
  summarise(total_countries = n_distinct(country))
```
```{r}
life_expectancy_tidy %>%
  summarise(total_countries = n_distinct(country))
```


## Life Expectancy  

3. (2 points) Let's limit the data to 100 years (1920-2020). For these years, which country has the highest average life expectancy? How about the lowest average life expectancy?  

*Sweden has the highest average life expectancy.* 
```{r}
life_expectancy_tidy %>%
  filter(year >= 1920 & year <= 2020) %>%
  group_by(country) %>%
  summarise(average_life_expectancy = mean(life_expectancy)) %>%
  arrange(desc(average_life_expectancy))
```
*Central African Republic has lowest average life expectancy.*
```{r}
life_expectancy_tidy %>%
  filter(year >= 1920, year <= 2020) %>%
  group_by(country) %>%
  summarise(average_life_expectancy = mean(life_expectancy)) %>%
  arrange(average_life_expectancy)
```


4. (3 points) Although we can see which country has the highest life expectancy for the past 100 years, we don't know which countries have changed the most. What are the top 5 countries that have experienced the biggest improvement in life expectancy between 1920-2020?

*Kuwait, Kyrgyz Republic, Turkmenistan, South Korea, Tajikistan are the top 5 countries that have experienced the biggest improvement in life expectancy between 1920-2020.*
```{r}
life_expectancy_tidy %>%
  filter(year == 1920 | year == 2020) %>%
  pivot_wider(
    names_from = "year",
    names_glue = "yr_{year}",
    values_from = "life_expectancy") %>%
  mutate(improved_life_expectancy = yr_2020 - yr_1920) %>%
  arrange(desc(improved_life_expectancy))
```

5. (3 points) Make a plot that shows the change over the past 100 years for the country with the biggest improvement in life expectancy. Be sure to add appropriate aesthetics to make the plot clean and clear. Once you have made the plot, do a little internet searching and see if you can discover what historical event may have contributed to this remarkable change.  

*Kuwait began the development of its petroleum industry in the 1930s. The industry is largely responsible for the country’s wealth and high standard of living today. The biggest increase in life expectancy was in the 1960s. Life expectancy in Kuwait grew rapidly with the coinciding economic development during the decade.*
```{r}
life_expectancy_tidy %>% 
  filter(country=="Kuwait") %>%
  filter(between(year, 1920, 2020)) %>% 
  ggplot(aes(x=year, y=life_expectancy, fill=country))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  theme(plot.title = element_text(size = rel(1.5), face='bold', hjust = 0.5))+
  theme(axis.text = element_text(size = 4))+
  labs(title = "Kuwait Life Expectancy From 1920 to 2020",
       x = "Year",
       y = "Life Expectancy")
```

## Population Growth
6. (3 points) Which 5 countries have had the highest population growth over the past 100 years (1920-2020)?  

*India, China, Indonesia, US, Pakistan have had the highest population growth over the past 100 years.*
```{r}
population_tidy %>% 
  filter(year == 1920 | year == 2020) %>% 
  pivot_wider(names_from = "year",
              names_glue = "yr_{year}",
              values_from = "population") %>% 
  mutate(pgrowth = yr_2020 - yr_1920) %>% 
  arrange(desc(pgrowth))
```

7. (4 points) Produce a plot that shows the 5 countries that have had the highest population growth over the past 100 years (1920-2020). Which countries appear to have had exponential growth?  

*China and India appear to have had exponential growth.*
```{r}
population_tidy %>% 
  filter(country %in% c("India","China","Indonesia","United States","Pakistan")) %>%
  filter(between(year, 1920, 2020)) %>% 
  ggplot(aes(x=year, y=population, group=country , color=country))+
  geom_line()+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size=4))+
  theme(plot.title = element_text(size = rel(1.5), face='bold', hjust = 0.5))+
  labs(title = "Population Growth over 100 Years",
       x = "Year",
       y = "Population")
  

```

## Income
The units used for income are gross domestic product per person adjusted for differences in purchasing power in international dollars.

8. (4 points) As in the previous questions, which countries have experienced the biggest growth in per person GDP. Show this as a table and then plot the changes for the top 5 countries. With a bit of research, you should be able to explain the dramatic downturns of the wealthiest economies that occurred during the 1980's.  

*Qatar, Luxembourg, Singapore, Brunei, Ireland have experienced the biggest growth in per person GDP. Between 1980 and 1982 the economy experienced a deep recession, caused by the Iranian Revolution which caused a disruption to the global oil supply. This affected many countries world wide.*
```{r}
income_per_person_tidy %>% 
  filter(year == 1920 | year == 2020) %>% 
  pivot_wider(names_from = "year",
              names_glue = "yr_{year}",
              values_from = "income") %>% 
  mutate(income_growth = yr_2020 - yr_1920) %>% 
  arrange(desc(income_growth))
```

```{r}
income_per_person_tidy %>% 
  filter(country %in% c("Qatar","Luxembourg","Singapore","Brunei","Ireland")) %>%
  filter(between(year, 1920, 2020)) %>% 
  ggplot(aes(x=year, y=income, group=country , color=country))+
  geom_line()+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size=4))+
  theme(plot.title = element_text(size = rel(1.5), face='bold', hjust = 0.5))+
  labs(title = "GDP Growth over 100 Years",
       x = "Year",
       y = "GDP growth")

```


9. (3 points) Create three new objects that restrict each data set (life expectancy, population, income) to the years 1920-2020. Hint: I suggest doing this with the long form of your data. Once this is done, merge all three data sets using the code I provide below. You may need to adjust the code depending on how you have named your objects. I called mine `life_expectancy_100`, `population_100`, and `income_100`. For some of you, learning these `joins` will be important for your project.  

```{r}
life_expectancy_100 <- 
  life_expectancy_tidy %>% 
  filter(between(year, 1920, 2020))
```

```{r}
population_100 <- 
  population_tidy %>% 
  filter(between(year, 1920, 2020))
```

```{r}
income_100 <- 
  income_per_person_tidy %>% 
  filter(between(year, 1920, 2020))
```

```{r}
gapminder_join <- inner_join(life_expectancy_100, population_100, by= c("country", "year"))
gapminder_join <- inner_join(gapminder_join, income_100, by= c("country", "year"))
gapminder_join
```

10. (4 points) Use the joined data to perform an analysis of your choice. The analysis should include a comparison between two or more of the variables `life_expectancy`, `population`, or `income.`  

**Make two graphs compare GDP per person, life expectancy, and population growth between China and the US. Notice in GDP growth of China, what happened in 1990s causing the growth shape? What about in the 1960s, what events drove the life expectancy to change? Finally compare China with US in all three graphs. **

```{r}
gapminder_join %>% 
  filter(country %in% c("China","United States")) %>%
  filter(between(year, 1920, 2020)) %>% 
  ggplot(aes(x=year, y=income, group=country , color=country))+
  geom_line()+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size=4))+
  theme(plot.title = element_text(size = rel(1.5), face='bold', hjust = 0.5))+
  labs(title = "US & China GDP Growth over 100 Years",
       x = "Year",
       y = "GDP growth")
```

*China adapted with some major policy changes with leadership, military, as well as social order, as Chairman Deng Xiao- Ping started to push reform. Many notable success in foreign relations also helped the economical growth in its 1990s.*

```{r}
gapminder_join %>% 
  filter(country %in% c("China","United States")) %>%
  filter(between(year, 1920, 2020)) %>% 
  ggplot(aes(x=year, y=life_expectancy, group=country , color=country))+
  geom_line()+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size=4))+
  theme(plot.title = element_text(size = rel(1.5), face='bold', hjust = 0.5))+
  labs(title = "US & China Life Expectancy over 100 Years",
       x = "Year",
       y = "Life Expectancy")
```

*The sudden drop in the life expectancy in the 1960s in China was likely due to the Great Chinese Famine, majorly contributed by inefficient distribution of food within the nation's planned economy, requiring the use of poor agricultural techniques, over-reporting of grain production, and ordering millions of farmers to switch to iron and steel production. Causing tens of millions of people to starvation and death.*

```{r}
gapminder_join %>% 
  filter(country %in% c("China","United States")) %>%
  filter(between(year, 1920, 2020)) %>% 
  ggplot(aes(x=year, y=population, group=country , color=country))+
  geom_line()+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size=4))+
  theme(plot.title = element_text(size = rel(1.5), face='bold', hjust = 0.5))+
  labs(title = "US & China Population Growth over 100 Years",
       x = "Year",
       y = "Population Growth")
```
*Looking at the above three graphs, we can observe that GDP per person in the US is in constant growing trend since the 1930s, whereas the curve of China started to show significant up-trend starting in the 1990s. Similarly, the life expectancy curve of the US showing less oscillation comparing to China, especially during the 1940s, and 1960s period. Comparing both countries' population growth curves, we can see even both countries are experiencing positive growth, but the Chinese population are growing exponentially since the 1950s. Looking at both countries, across these three aspects, US have more steady and predictable curves than China, and the growth curves of China, might need more years to reflect the country's true economy.*




